---
title: "Project 1 - Exploring the Determinants of Spotify Track Popularity: Audio Features, Genre, and Explicit Content"
author: "Ruofei(Sophia) Du, Meghan Phillips, Harivinay Gujjula"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
library(dplyr)
library(ggplot2)
```

```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(results="markup", warning = F, message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

#1. Introduction 

##1.1 About the Data Set

Our dataset was obtained from Kaggle by researching music-related dataset. The dataset contains more than 114,000 observations and 21 variables related to Spotify tracks. These variables include track information such as track name and artist name, the popularity of the songs, as well as several audio features, including energy, danceability, valence, tempo, and other characteristics that describe the musical properties of each song.

```{r}
spo <- read.csv("spotifydataset.csv")
str(spo)
summary(spo)
dim(spo)
```
##1.2 Research Questions

There are many possible ways to explore this dataset. In this project, we are particularly interested in understanding what makes a song popular.

First, we explore whether genre plays a role in song popularity. Since artist popularity may influence the popularity of a track, we remove the artist variable from our analysis to focus only on the audio features of the songs.

This leads to several research questions:

Do audio features such as danceability, energy, tempo, and valence influence the popularity of a song?

Does genre affect song popularity?

If audio features do not directly affect popularity, do they influence how a song is categorized into genres, which in turn affects popularity? In that case, genre may act as a confounding variable between audio features and popularity.

Does the presence of explicit content make a song more or less popular?

#2 Data Preparation 

##2.1 Data Cleaning

To investigate these research questions, we first clean the dataset and remove columns that are not relevant to our analysis.

For example, the tempo (BPM) of a song cannot realistically be 0. Although this value is not recorded as NA in the dataset, it still represents an invalid observation and should be removed during the data cleaning process.

Based on the information provided in the dataset, we focus on variables that may influence song popularity. In particular, we examine whether audio features or genre have any effect on the popularity of songs. Therefore, we select the relevant variables, including popularity, several audio features (danceability, energy, valence, and tempo), the explicit content indicator, and the track genre.

```{r}
spo1 <- spo[, c(
  "popularity",
  "danceability",
  "energy",
  "valence",
  "tempo",
  "explicit",
  "track_genre"
)]
str(spo1)
summary(spo1)
colSums(is.na(spo1))
```
```{r}
spo1$explicit <- as.factor(spo1$explicit)
str(spo1)
```
Furthur cleaning the data, as the BPM can't be 0, but the summary showed that the min(tempo) = 0.
```{r}
sum(spo1$tempo == 0)
```
We can see that there are still 157 rows with tempo = 0, which is inproper cause we can't have a song with BPM = 0. We get rid of those missing data.

```{r}
spo1 <- spo1[spo1$tempo > 0, ]
sum(spo1$tempo == 0)
write.csv(spo1,"cleaned_spodata.csv")
```
##2.2 Further Data Transformation
To analyze the relationship between genre and audio features or popularity more effectively, the original dataset contains too many genres (over 120) to interpret clearly. Therefore, we manually categorized all genres into four broader genre families based on common musical classifications and similarities in musical style. Each family is expected to share similar musical characteristics and, to some extent, similar audio feature patterns.

To ensure smmothy analysis, we make sure that the family column was properly turned into factors as it’s a categorical variable. 

```{r}
spo1$family <- case_when(
  spo1$track_genre %in% c("rock","alt-rock","hard-rock","metal",
    "heavy-metal","death-metal","black-metal","metalcore","grunge",
    "punk","punk-rock","emo","psych-rock","goth","hardcore",
    "grindcore","rock-n-roll","rockabilly","j-rock","alternative",
    "british","guitar","indie") ~ "Rock/Metal",
  
  spo1$track_genre %in% c("edm","electronic","electro","house",
    "deep-house","chicago-house","progressive-house","techno","trance",
    "dubstep","drum-and-bass","garage","disco","ambient","chill",
    "idm","minimal-techno","detroit-techno","breakbeat","hardstyle",
    "industrial","club","dance","trip-hop","happy","party",
    "hip-hop","r-n-b","funk","soul","groove") ~ "Electronic/Dance",
  
  spo1$track_genre %in% c("latin","latino","reggaeton","salsa",
    "samba","brazil","mpb","pagode","sertanejo","forro","tango",
    "spanish","afrobeat","world-music","turkish","iranian","malay",
    "indian","french","german","dancehall","reggae","ska",
    "dub","romance",
    "pop","k-pop","j-pop","pop-film","power-pop","synth-pop",
    "cantopop","mandopop","indie-pop","j-idol","j-dance",
    "swedish","anime","disney") ~ "Latin/Pop",
  
  TRUE ~ "Acoustic/Classical"
)
spo1$family <- as.factor(spo1$family)
table(spo1$family)
```


#Section 3 Exploratory and Statistical Data Analysis
This section starts with exploring the relationships among variables through visualizations and descriptive analysis, followed by simple statistical tests to evaluate whether the observed patterns worth further analysis.
##3.1 Genre Families and Audio Feature Patterns
To better understand the relationship between genre families and audio features, we first examine how different genres are distributed across the audio-feature space. Since four audio feature variables are considered in this study (danceability, energy, valence, and tempo), we visualize their relationships using two scatterplots. 
```{r}
genre_profile <- spo1 %>%
  group_by(track_genre) %>%
  summarise(
    danceability = mean(danceability),
    energy = mean(energy),
    valence = mean(valence),
    tempo = mean(tempo),
    n = n(),
    .groups = "drop"
  )
genre_family <- spo1 %>%
  select(track_genre, family) %>%
  distinct()

genre_profile <- genre_profile %>%
  left_join(genre_family, by = "track_genre")
ggplot(genre_profile,
       aes(x = energy, y = danceability, color = family)) +
  geom_point(size = 3, alpha = 0.8) +
stat_ellipse(aes(group = family), linewidth = 0.8) +
  labs(title = "Audio Features Distribution by Genre Family: Energy and Danceability",
       x = "Energy (mean)", y = "Danceability (mean)", color = "Family") +
  theme_minimal()
```
The first plot shows the relationship between **energy** and **danceability** across genre families. The scatterplot illustrates how different genres are distributed across the energy and danceability feature space. Each point represents the mean audio feature values for a specific genre. The ellipses highlight the approximate distribution regions of the four genre families.
```{r}
ggplot(genre_profile,
       aes(x = valence, y = tempo, color = family)) +
  geom_point(size = 3, alpha = 0.8) +
stat_ellipse(aes(group = family), linewidth = 0.8) +
  labs(title = "Audio Features Distribution by Genre Family: Valence and Tempo",
       x = "Valence (mean)", y = "Tempo (mean)", color = "Family") +
  theme_minimal()
```
The second plot visualizes the relationship between **valence** and **tempo** across genre families. Compared with the first scatterplot (**energy** and **danceability**), the clusters among genre families overlap substantially and appear less distinct. Although some tendencies can still be observed, such as relatively higher tempo values for **Electronic/Dance** family and lower valence for **Acoustic/Classical** family, the overlap among families is much stronger than in the energy/danceability feature space. This suggests that **valence** and **tempo** alone are less effective at differentiating genres than **energy** and **danceability**.
Therefore, based on these visualizations, **energy** and **danceability** are expected to play a stronger role in distinguishing musical styles across genre families.

##3.2 Distribution of Popularity and Audio Features

Before analyzing the relationships between variables, it is useful to first understand the overall distribution of popularity and the 4 audio features in the dataset. Examining the distributions helps identify general patterns, potential skewness, and the range of values present in the data.

The following 5 histograms illustrate the distributions of track popularity and four key audio features: danceability, energy, valence, and tempo. 

```{r}
library(ggplot2)
ggplot(spo1, aes(x = popularity)) +
  geom_histogram(fill = "skyblue", bins = 30, color = "white") +
  labs(
    title = "Distribution of Track Popularity",
    x = "Popularity Score",
    y = "Count"
  ) +
  theme_minimal()
sum(spo1$popularity == 0)
```
The distribution of the popularity is highly right-skewed. A large number of very low popularity scores were observed on the left-hand side near zero. As popularity increases, the number of tracks gradually decreases. The summary showed that we have over 16000 zero values, which is more than 10% of the data set. 
The skewed distribution may influence subsequent analysis. A large portion of the tracks have extremely low popularity values, which might result in a weak correlation between audio features and popularity. To better understand and reduce these effects, the group popularity was sliced into five ranges to compare the audio features across these popularity groups.

```{r}
ggplot(spo1, aes(x = danceability)) +
  geom_histogram(fill = "darkgreen", bins = 30, color = "white") +
  labs(
    title = "Distribution of Danceability",
    x = "Danceability",
    y = "Count"
  ) +
  theme_minimal()
```
The distribution of danceability appears an approximate bell-shaped, with most tracks concentrated between 0.4 and 0.7 danceability. This suggests that a large majority of the dataset have a moderate level of danceability, while the songs with extremely high or extreme low danceability may be relatively rare. 
This symmetric distribution shows that the danceability is relatively balanced among the dataset, which could result in an easier comparison for the danceability with different popularity ranges. 

```{r}
ggplot(spo1, aes(x = energy)) +
  geom_histogram(fill = "salmon", bins = 30, color = "white") +
  labs(
    title = "Distribution of Energy",
    x = "Energy",
    y = "Count"
  ) +
  theme_minimal()
```
The distribution of energy is different from danceability. It is every left-skewed, indicating that there are large numbers of songs with higher energy comparing to lower energy. The majority of the songs are concentrated between approximately 0.6 and 0.9. The lower the energy, the lesser the frequency. 
This suggests that the songs in the dataset tend to be energetic, which is consistent with the characteristics of morden music, where majority of the songs produced are lounder, faster and more dynamic. The soft and calmer songs may represents a relative smaller portion in the dataset. 
```{r}
ggplot(spo1, aes(x = valence)) +
  geom_histogram(fill = "violet", bins = 30, color = "white") +
  labs(
    title = "Distribution of Valence",
    x = "Valence",
    y = "Count"
  ) +
  theme_minimal()
```
The valence in spotify means that the higher the valence is, the song is more likely to be happier and positive. The distribution of valence is relatively spread out very average aross full range between 0 and 1, without a strong skew toward either low or high valence. A slightly larger number of songs appear in mid-range valence, but overall the distribution is very even. 

The pattern suggests that the emotional tone of songs varies a lot across the dataset. This indicates that he emotional tone is very diverse across the dataset. 

```{r}
ggplot(spo1, aes(x = tempo)) +
  geom_histogram(fill = "orange", bins = 30, color = "white") +
  labs(
    title = "Distribution of Tempo",
    x = "Tempo (BPM)",
    y = "Count"
  ) +
  theme_minimal()
```
The distribution of tempo shows that most tracks are concentrated between 90 - 140 BPM, with the highest density around a range of 110 - 130 BPM. The very slow tempos (< 70 BPM) and extremely fast tempos (< 170 BPM) are very rare in the dataset. 
The patterns suggest that the tempos of the songs in this dataset fall within a moderate tempo range, which matches the common characteristics of many genres of modern pop music. The very fast tracks and very slow tracks are less common. 

```{r}
ggplot(spo1, aes(x = danceability, y = popularity)) +
  geom_point(color = "darkgreen", alpha = 0.5) +
  labs(
    title = "Popularity vs Danceability",
    x = "Danceability",
    y = "Popularity"
  ) +
  theme_minimal()
```
The scatterplot of popularity versus danceability does not show a clear linear relationship. The points scattered across all range of danceability, suggesting that the songs with both low and high danceability could achieve a wide range of popularity scores. 
A slight positive correlation was found between danceability and popularity, but the correlation is very weak. This indicates that the danceability alone is unlikely to be a strong predictor of song popularity. 
```{r}
ggplot(spo1, aes(x = energy, y = popularity)) +
  geom_point(color = "salmon", alpha = 0.5) +
  labs(
    title = "Popularity vs Danceability",
    x = "Energy",
    y = "Popularity"
  ) +
  theme_minimal()
```
The scatterplot of popularity versus energy shows an even weaker correlation than danceability and popularity. Songs with both high and low energy levels appear across the entire range of popularity. Although slightly more songs with high energy concentrated in the higher popularity range, the correlation is very weak. This suggest that the energy alone is not a strong predictor of song popularity. 
```{r}
ggplot(spo1, aes(x = valence, y = popularity)) +
  geom_point(color = "violet", alpha = 0.5) +
  labs(
    title = "Popularity vs Valence",
    x = "Valence",
    y = "Popularity"
  ) +
  theme_minimal()
```
Similarly, the relationship between valence and popularity appears weak. Songs with both positive and negative emotional tones distributed across a wide popularity range. The scatterplot suggests that the emotional tone is not likely to be a strong predictor for a song’s popularity. 
```{r}
ggplot(spo1, aes(x = tempo, y = popularity)) +
  geom_point(color = "orange", alpha = 0.5) +
  labs(
    title = "Popularity vs Tempo",
    x = "Tempo",
    y = "Popularity"
  ) +
  theme_minimal()
```
The scatterplot of popularity versus tempo also shows no obvious relationship. Most songs fall within the moderate tempo range observed earlier, yet their popularity scores still vary substantially. This indicates that tempo alone does not have a strong influence on the popularity of a track.

Overall, the scatterplots suggest none of the individual audio features show a strong direct relationship with popularity. Within all four audio features, only danceability shows a relatively stronger.
```{r}
cor(spo1$danceability, spo1$popularity)
cor(spo1$energy, spo1$popularity)
cor(spo1$valence, spo1$popularity)
cor(spo1$tempo, spo1$popularity)
```
To qualify these relationships, correlation coefficients were calculated. The result show that all correlations are very weak: danceability (0.0365), energy (0.0016), valence (-0.0402), and tempo (0.0144). These values are extremely weak linear relationships between the audio features and the popularity. This suggests that the song popularity is likely to be influenced by other factors beyond audio characteristics, such as artist popularity, broader genre trends, etc. 

##3.3 Relationship between Popularity Groups and Audio Features 

In the pervious part, the scatterplots did not reveal a clear relationship between four individual audio features and popularity, we further explore the data by grouping the popularity into different levels. This approach allow us to visualize whether songs with different popularity levels tend to have different audio characteristics. 

```{r}
spo1$pop_group <- cut(
  spo1$popularity,
  breaks = c(0, 20, 40, 60, 80, 100),
  labels = c("Very Low", "Low", "Medium", "High", "Very High"),
  include.lowest = TRUE
)

spo1$pop_group <- as.factor(spo1$pop_group)
table(spo1$pop_group)
```
Popularity scores are divided into five different groups: Very Low (0-20), Low (20-40), Medium(40-60), High(60-80) and Very High(80-100). Now we can visualize how audio features vary across different levels of song popularity. 

```{r}
ggplot(spo1, aes(x = pop_group, y = danceability, fill = pop_group)) +
  geom_boxplot() +
  labs(
    title = "Danceability across Popularity Groups",
    x = "Popularity Group",
    y = "Danceability"
  ) +
  theme_minimal() +
  guides(fill = "none")
ggplot(spo1, aes(x = pop_group, y = energy, fill = pop_group)) +
  geom_boxplot() +
  labs(
    title = "Energy across Energy Groups",
    x = "Popularity Group",
    y = "Energy"
  ) +
  theme_minimal() +
  guides(fill = "none")
ggplot(spo1, aes(x = pop_group, y = valence, fill = pop_group)) +
  geom_boxplot() +
  labs(
    title = "Valence across Popularity Groups",
    x = "Popularity Group",
    y = "Valence"
  ) +
  theme_minimal() +
  guides(fill = "none")
ggplot(spo1, aes(x = pop_group, y = tempo, fill = pop_group)) +
  geom_boxplot() +
  labs(
    title = "Tempo across Popularity Groups",
    x = "Popularity Group",
    y = "Tempo"
  ) +
  theme_minimal() +
  guides(fill = "none")
```
The boxplot for danceability shows a slightly increase in median values as popularity increases. Songs in “Very High” popularity group tend to have slightly higher danceability compared to tracks with less popularity.  However, the distribution still overlaps substantially across all groups, suggesting that the effect is relatively weak. 
The distribution of energy across different popularity groups are very similar. Median values and interquartile ranges overlap heavily, suggesting that the energy level does not differ significantly between more popular and less popular songs. 

Valence also shows very little differences across popularity groups, the median values is close to each other across all popularity levels, suggesting that valence is not a strong influence on songs’ popularity. 

Similarly, tempo share a similar distribution with valence and energy. The tempo distribution are nearly identical across popularity levels. Most songs fall within the same tempo range regardless of their popularity, suggesting that tempo is anot a strong influence on popularity. 

Overall, the boxplots suggests that audio features only slightly influence the popularity of a songs. While danceability appears slightly higher for the most popular songs, the distribution of valence, energy and tempo remain largely similar across all popularity groups. This indicates that the individual audio features are unlikely to be a strong influence on popularity. Danceability is the strongest feature to affect the popularity among the four audio features. 

To statistically evaluate whether audio features differ across popularity groups, one-way ANOVA test were conducted for each feature.

```{r}
anova_dance <- aov(danceability ~ pop_group, data = spo1)
summary(anova_dance)
```
```{r}
get_eta_sq <- function(feature_name, data) {
  formula <- as.formula(paste(feature_name, "~ pop_group"))
  model <- aov(formula, data = data)
  tab <- summary(model)[[1]]
  
  SS_between <- tab["pop_group", "Sum Sq"]
  SS_total   <- sum(tab[, "Sum Sq"])
  
  eta_sq <- SS_between / SS_total
  
  return(list(
    feature = feature_name,
    eta_sq = eta_sq,
    p_value = tab["pop_group", "Pr(>F)"]
  ))
}
results <- rbind(
  unlist(get_eta_sq("danceability", spo1)),
  unlist(get_eta_sq("energy", spo1)),
  unlist(get_eta_sq("valence", spo1)),
  unlist(get_eta_sq("tempo", spo1))
)

results <- as.data.frame(results)
results
```
The p value of all four audio features vs popularity levels are extremely small, indicating that all four audio features show statistically significant differences across popularity levels. However, the effect sizes (η²) are relatively small.
Among the features, danceability exhibits the strongest relationship with popularity groups (F = 292.1, η² = 0.0102), followed by valence (F = 166.1, η² = 0.0058), energy (F = 97.8, η² = 0.0034), and tempo (F = 29.4, η² = 0.0010).
Despite statistical significance, the small η² values indicate that popularity groups explain only a very small portion of the variance in these audio characteristics(~1%). This suggests that although certain audio features such as danceability may slightly differ across popularity levels, audio features alone are unlikely to be strong determinants of song popularity.

## 3.4 Genre Families and Popularity

In addition to audio features, genre may be a factor to influence how popular a song is. Different genres often reflect distinct musical styles and audience preferences, which may affect listener engagement and streaming performance. Therefore, we further examine whether song popularity differs across the four genre families defined earlier.

```{r}
anova_pop_fam <- aov(popularity ~ family, data = spo1)
summary(anova_pop_fam)
tab_pop <- summary(anova_pop_fam)[[1]]
eta_sq_pop <- tab_pop[1, "Sum Sq"] / sum(tab_pop[, "Sum Sq"])
cat(sprintf("Eta squared (η²) = %.4f\n", eta_sq_pop))
```
The ANOVA result show that popularity differs across different genre families. (F = 237, p < 0.001). However, the effect size is relatively small (η² = 0.0062), indicating that genre explains only a small proportion of the variance in song popularity.

```{r}
spo1 %>%
  group_by(family) %>%
  summarise(mean_popularity = mean(popularity)) %>%
  arrange(desc(mean_popularity))
```
The average popularity scores show that Rock/Metal and Latin/Pop genres have slightly higher popularity levels, while Acoustic/Classical songs tend to have lower popularity scores.

```{r}
spo1 %>%
  group_by(family) %>%
  summarise(
    danceability = mean(danceability),
    energy       = mean(energy),
    valence      = mean(valence),
    tempo        = mean(tempo)
  )
```
```{r}
get_eta_sq_fam <- function(model) {
  tab <- summary(model)[[1]]
  round(tab[1, "Sum Sq"] / sum(tab[, "Sum Sq"]), 4)
}
anova_dance_fam   <- aov(danceability ~ family, data = spo1)
anova_energy_fam  <- aov(energy       ~ family, data = spo1)
anova_valence_fam <- aov(valence      ~ family, data = spo1)
anova_tempo_fam   <- aov(tempo        ~ family, data = spo1)

summary(anova_dance_fam)
summary(anova_energy_fam)
summary(anova_valence_fam)
summary(anova_tempo_fam)

cat("Eta squared:\n")
cat(sprintf("  Danceability: %.4f\n", get_eta_sq_fam(anova_dance_fam)))
cat(sprintf("  Energy:       %.4f\n", get_eta_sq_fam(anova_energy_fam)))
cat(sprintf("  Valence:      %.4f\n", get_eta_sq_fam(anova_valence_fam)))
cat(sprintf("  Tempo:        %.4f\n", get_eta_sq_fam(anova_tempo_fam)))
```
Further analysis reveals that genre families strongly differ in their audio characteristics. For example, Electronic/Dance songs exhibit higher danceability and energy, while Rock/Metal songs show the highest energy but lower danceability. The effect sizes for these differences are substantially larger, particularly for energy (η² = 0.163) and danceability (η² = 0.087).

However, the actual means are not as strongly different as expected statistically. This is because the η² is so small that the result only represents a small portion of the variance. 

##3.5 Relationship between Audio Features and Popularities By Genre Families

To further visualize how the relationship between audio features and popularity may vary across different music styles, scatterplots colored by genre family are shown. 

```{r}
ggplot(spo1,
       aes(x = danceability,
           y = popularity,
           color = family)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Popularity vs Danceability by Music Family",
    x = "Danceability",
    y = "Popularity",
    color = "Family"
  ) +
  theme_minimal()

ggplot(spo1,
       aes(x = energy,
           y = popularity,
           color = family)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Popularity vs Energy by Music Family",
    x = "Energy",
    y = "Popularity",
    color = "Family"
  ) +
  theme_minimal()

ggplot(spo1,
       aes(x = valence,
           y = popularity,
           color = family)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Popularity vs Valence by Music Family",
    x = "Valence",
    y = "Popularity",
    color = "Family"
  ) +
  theme_minimal()

ggplot(spo1,
       aes(x = tempo,
           y = popularity,
           color = family)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Popularity vs Tempo by Music Family",
    x = "Tempo",
    y = "Popularity",
    color = "Family"
  ) +
  theme_minimal()
```
The scatterplots colored by genre family provide additional insights into the relationship between audio features and popularity. Although different genre families occupy slightly different regions of the audio feature space, the relationship between audio features and popularity remains weak across all genres.
The points are widely dispersed with heavy overlap between genre families, indicating that songs from different genres can have similar popularity levels across a wide range of audio feature values.
These visual patterns are consistent with the earlier statistical results, suggesting that neither individual audio features nor genre families alone strongly statistically determine song popularity. This is a result that within expectation, because the dataset is extremely large. 




Can we see any differences in Genre when we look at Tempo x Popularity? There are too many points! A regular scatterplot was done, coloring each point by genre. Since there are over 100 genres, it was difficult to see anything. Plus, it appears that the scatterplot added layers to the dots, instead of some blend. This helped us with our ideas about clustering for project two.

```{r}
ggplot(spo1, aes(x = tempo, y = popularity, color= track_genre)) + geom_point(alpha = 0.5)
```
What is another way to look at plots by Genre? Possibly in facets! This allows us to see the tempo by popularity for each individual genre at a higher level, showing differences between genres. 
```{r}
ggplot(spo1, aes(x = tempo, y = popularity)) + geom_point(color= "blue", alpha = 0.3) + facet_wrap(~ track_genre) 
```

Getting to our EDA question, Are explicit tracks more or less popular than non-explicit tracks?
we want to do a t-test to check. This showed comparison of average popularity between explicit and non-explicit tracks. Explicit songs had a higher mean popularity (36.5) compared to non-explicit songs (32.9). The difference was statistically significant (t = -14, with p value of <2e-16. The 95% confidence interval for the difference in means was (-4.02, -3.02), which showed that  explicit tracks are on average between 3.02 and 4.02 popularity points higher than non-explicit tracks. While this could suggest that explicit content is associated with higher popularity, we want to do some more test to get a better picture because it doesn't seem like that is a very big difference in popularity score and we must remember that this is such a large sample size, as well as the fact that statistically significant does not mean correlation.
```{r}
t.test(popularity ~ explicit, data = spo1)
```

```{r}
lower <- quantile(spo1$popularity, 0.25, na.rm = TRUE)
upper <- quantile(spo1$popularity, 0.75, na.rm = TRUE)
spo1_top_pop<- subset(spo1, popularity >= upper)
t.test(popularity ~ explicit, data = spo1_top_pop)
spo1_middle_pop <- subset(spo1, popularity > lower & popularity <= upper)
t.test(popularity ~ explicit, data = spo1_middle_pop)
spo1_bottom_pop <- subset(spo1, popularity < lower)
t.test(popularity ~ explicit, data = spo1_bottom_pop)
```

#This splits the most popular songs into sections based on popularity and does a t test on the groups in order to compare means in explicit vs non-explicit tracks. Mean popularity for explicit tracks were not higher throughout the different all levels of popularity. This lead to another question of wanting to look at the top 10% most popular songs and top 5%. Popularity was slightly more in these. Then we wanted to run correlation coefficient to really know.
```{r}
cut_pop90<- quantile(spo1$popularity, 0.90)
spo1_pop90 <- subset(spo1, popularity > cut_pop90)
cut_pop95<- quantile(spo1$popularity, 0.95)
spo1_pop95 <- subset(spo1, popularity > cut_pop95)
t.test(popularity ~ explicit, data = spo1_pop90)
t.test(popularity ~ explicit, data = spo1_pop95)
```

#Correlation between explicit and popularity- shows extremely low (r = 0.044). This indicates that while the difference is statistically detectable due to the large sample size, whether a track is explicit explains very little of the overall variation in popularity. Therefore we must note the difference between a statistical significance versus very very low correllation. 

```{r}
spo1$explicit_num<- as.numeric(spo1$explicit)
cor(spo1$popularity, spo1$explicit_num)
```


#Splitting popularity by values 
```{r}
pop_20df <- spo1 %>%
  filter(popularity >= 0 & popularity <= 20)
pop_40df <- spo1 %>%
  filter(popularity >= 20 & popularity <= 40)
pop_60df <- spo1 %>%
  filter(popularity >= 40 & popularity <= 60)
pop_80df <- spo1 %>%
  filter(popularity >= 60 & popularity <= 80)
pop_100df <- spo1 %>%
  filter(popularity >= 80 & popularity <= 100)
```
#Calculating correlation values of audio features(danceability, energy, tempo and valence) to popularity for each group

```{r}
features <- c("danceability", "energy", "valence", "tempo", "popularity")

cor_pop_20 <- cor(pop_20df[, features], pop_20df$popularity,
use = "complete.obs")
cor_pop_40 <- cor(pop_40df[, features], pop_40df$popularity,
use = "complete.obs")
cor_pop_60 <- cor(pop_60df[, features], pop_60df$popularity,
use = "complete.obs")
cor_pop_80 <- cor(pop_80df[, features], pop_80df$popularity,
use = "complete.obs")
cor_pop_100 <- cor(pop_100df[, features], pop_100df$popularity,
use = "complete.obs")
```

#Plotting heatmap of correlation values

```{r}
library(reshape2)
library(ggplot2)

plot_corr <- function(cor_matrix, title_text, filename) {
  
  melted <- melt(cor_matrix)
  
  p <- ggplot(melted, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    geom_text(aes(label = round(value, 2)),
              size = 4) +
    scale_fill_gradient2(low = "blue",
                         mid = "white",
                         high = "red",
                         midpoint = 0,
                         limits = c(-1,1)) +
    theme_minimal() +
    labs(title = title_text,
         x = "",
         y = "",
         fill = "Correlation")
  
  ggsave(filename, plot = p, width = 6, height = 5, dpi = 300)
  
  return(p)
}


plot_corr(cor_pop_20,
          "Correlation Matrix: Popularity 0-20 Range", "corr_pop_20.png")
plot_corr(cor_pop_40,
          "Correlation Matrix: Popularity 20-40 Range", "corr_pop_40.png")
plot_corr(cor_pop_60,
          "Correlation Matrix: Popularity 40-60 Range", "corr_pop_60.png")
plot_corr(cor_pop_80,
          "Correlation Matrix: Popularity 60-80 Range", "corr_pop_80.png")
plot_corr(cor_pop_100,
          "Correlation Matrix: Popularity 80-100 Range", "corr_pop_100.png")
```

#Converting correlation values from wide format into the long format to compare each audio feature across different popularity groups
```{r}
features <- c("danceability", "energy", "valence", "tempo")

df_list <- list(
  "0-20"   = pop_20df,
  "20-40"  = pop_40df,
  "40-60"  = pop_60df,
  "60-80"  = pop_80df,
  "80-100" = pop_100df
)

cor_long <- imap_dfr(df_list, ~{
  d <- .x
  tibble(
    Pop_Group = .y,
    Feature = features,
    Correlation = map_dbl(features, \(f) cor(d[[f]], d$popularity, use = "complete.obs"))
  )
})

cor_long$Pop_Group <- factor(cor_long$Pop_Group,
                             levels = c("0-20","20-40","40-60","60-80","80-100"))
```

#Plotting a line plot for correlation values between danceability and popularity across different popularity groups
```{r}
dance_cor <- cor_long %>% filter(Feature == "danceability")

p <- ggplot(dance_cor, aes(x = Correlation , y = Pop_Group, group = 1)) +
  geom_line() +
  geom_point(size = 3, color = "firebrick") +
  theme_minimal() +
  labs(title = "Correlation(popularity, danceability) across popularity groups",
       x = "Correlation", y = "Popularity group")

p
ggsave("danceability_corr_across_groups.png", p, width = 7, height = 4.5, dpi = 300)
```

#Plotting a line plot for correlation values between energy and popularity across different popularity groups

```{r}
energy_cor <- cor_long %>% filter(Feature == "energy")

p <- ggplot(energy_cor, aes(x = Correlation , y = Pop_Group, group = 1)) +
  geom_line() +
  geom_point(size = 3, color = "steelblue") +
  theme_minimal() +
  labs(title = "Correlation(popularity, energy) across popularity groups",
       x = "Correlation", y = "Popularity group")

p
ggsave("energy_corr_across_groups.png", p, width = 7, height = 4.5, dpi = 300)

```

#Plotting a line plot for correlation values between valence and popularity across different popularity groups

```{r}
valence_cor <- cor_long %>% filter(Feature == "valence")

p <- ggplot(valence_cor, aes(x = Correlation , y = Pop_Group, group = 1)) +
  geom_line() +
  geom_point(size = 3, color = "aquamarine") +
  theme_minimal() +
  labs(title = "Correlation(popularity, valence) across popularity groups",
       x = "Correlation", y = "Popularity group")

p
ggsave("valence_corr_across_groups.png", p, width = 7, height = 4.5, dpi = 300)
```

#Plotting a line plot for correaltion values between tempo and popularity across different popularity groups

```{r}
tempo_cor <- cor_long %>% filter(Feature == "tempo")

p <- ggplot(tempo_cor, aes(x = Correlation , y = Pop_Group, group = 1)) +
  geom_line() +
  geom_point(size = 3, color = "coral") +
  theme_minimal() +
  labs(title = "Correlation(popularity, tempo) across popularity groups",
       x = "Correlation", y = "Popularity group")

p
ggsave("tempo_corr_across_groups.png", p, width = 7, height = 4.5, dpi = 300)

```

Our general findings were that Popularity has no significant correlation with danceability, energy, tempo and valence, genres within the same music family share similar audio features, and the popularities differ for different families, and no correlation was found between explicit songs and popularity. 


