---
title: "Project 1"
author: "Ruofei(Sophia) Du, Meghan Phillips, Harivinay Gujjula"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
library(dplyr)
```

-Sophia
```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(results="markup", warning = F, message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

```{r}
spo <- read.csv("spotifydataset.csv")
str(spo)
summary(spo)
dim(spo)
```
Clean the data and get rid of the columns we don't need.
```{r}
spo1 <- spo[, c(
  "popularity",
  "danceability",
  "energy",
  "valence",
  "tempo",
  "explicit",
  "track_genre"
)]
str(spo1)
summary(spo1)
colSums(is.na(spo1))
```
```{r}
spo1$explicit <- as.factor(spo1$explicit)
str(spo1)
```
Furthur cleaning the data, as the BPM can't be 0, but the summary showed that the min(tempo) = 0.
```{r}
sum(spo1$tempo == 0)
```
We can see that there are still 157 rows with tempo = 0, which is inproper cause we can't have a song with BPM = 0. We get rid of those missing data.

```{r}
spo1 <- spo1[spo1$tempo > 0, ]
sum(spo1$tempo == 0)
write.csv(spo1,"cleaned_spodata.csv")
```

```{r}
spo1$family <- case_when(
  spo1$track_genre %in% c("rock","alt-rock","hard-rock","metal",
    "heavy-metal","death-metal","black-metal","metalcore","grunge",
    "punk","punk-rock","emo","psych-rock","goth","hardcore",
    "grindcore","rock-n-roll","rockabilly","j-rock","alternative",
    "british","guitar","indie") ~ "Rock/Metal",
  
  spo1$track_genre %in% c("edm","electronic","electro","house",
    "deep-house","chicago-house","progressive-house","techno","trance",
    "dubstep","drum-and-bass","garage","disco","ambient","chill",
    "idm","minimal-techno","detroit-techno","breakbeat","hardstyle",
    "industrial","club","dance","trip-hop","happy","party",
    "hip-hop","r-n-b","funk","soul","groove") ~ "Electronic/Dance",
  
  spo1$track_genre %in% c("latin","latino","reggaeton","salsa",
    "samba","brazil","mpb","pagode","sertanejo","forro","tango",
    "spanish","afrobeat","world-music","turkish","iranian","malay",
    "indian","french","german","dancehall","reggae","ska",
    "dub","romance",
    "pop","k-pop","j-pop","pop-film","power-pop","synth-pop",
    "cantopop","mandopop","indie-pop","j-idol","j-dance",
    "swedish","anime","disney") ~ "Latin/Pop",
  
  TRUE ~ "Acoustic/Classical"
)


spo1$family <- as.factor(spo1$family)
table(spo1$family)
```
Danceability vs Energy
```{r}
genre_profile <- spo1 %>%
  group_by(track_genre) %>%
  summarise(
    danceability = mean(danceability),
    energy = mean(energy),
    valence = mean(valence),
    tempo = mean(tempo),
    n = n(),
    .groups = "drop"
  )
genre_family <- spo1 %>%
  select(track_genre, family) %>%
  distinct()

genre_profile <- genre_profile %>%
  left_join(genre_family, by = "track_genre")
ggplot(genre_profile,
       aes(x = energy, y = danceability, color = family)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "Genre Distribution by Music Family: Energy vs Danceability",
       x = "Energy (mean)", y = "Danceability (mean)", color = "Family") +
  theme_minimal()


```
Valence vs Tempo
```{r}
ggplot(genre_profile,
       aes(x = valence, y = tempo, color = family)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "Genre Distribution by Music Family: Valence vs Tempo",
       x = "Valence (mean)", y = "Tempo (mean)", color = "Family") +
  theme_minimal()
```
```{r}

anova_pop_fam <- aov(popularity ~ family, data = spo1)
summary(anova_pop_fam)

tab_pop <- summary(anova_pop_fam)[[1]]
eta_sq_pop <- tab_pop[1, "Sum Sq"] / sum(tab_pop[, "Sum Sq"])
cat(sprintf("Eta squared (η²) = %.4f\n", eta_sq_pop))
```
```{r}
spo1 %>%
  group_by(family) %>%
  summarise(mean_popularity = mean(popularity)) %>%
  arrange(desc(mean_popularity))
```
```{r}
spo1 %>%
  group_by(family) %>%
  summarise(
    danceability = mean(danceability),
    energy       = mean(energy),
    valence      = mean(valence),
    tempo        = mean(tempo)
  )
```
```{r}
get_eta_sq_fam <- function(model) {
  tab <- summary(model)[[1]]
  round(tab[1, "Sum Sq"] / sum(tab[, "Sum Sq"]), 4)
}
anova_dance_fam   <- aov(danceability ~ family, data = spo1)
anova_energy_fam  <- aov(energy       ~ family, data = spo1)
anova_valence_fam <- aov(valence      ~ family, data = spo1)
anova_tempo_fam   <- aov(tempo        ~ family, data = spo1)

summary(anova_dance_fam)
summary(anova_energy_fam)
summary(anova_valence_fam)
summary(anova_tempo_fam)

cat("Eta squared:\n")
cat(sprintf("  Danceability: %.4f\n", get_eta_sq_fam(anova_dance_fam)))
cat(sprintf("  Energy:       %.4f\n", get_eta_sq_fam(anova_energy_fam)))
cat(sprintf("  Valence:      %.4f\n", get_eta_sq_fam(anova_valence_fam)))
cat(sprintf("  Tempo:        %.4f\n", get_eta_sq_fam(anova_tempo_fam)))
```

Q2
Now let's start with EDA. First, we plot some ggplots to observe the distribution. 
```{r}
library(ggplot2)
ggplot(spo1, aes(x = popularity)) +
  geom_histogram(fill = "skyblue", bins = 30, color = "white") +
  labs(
    title = "Distribution of Track Popularity",
    x = "Popularity Score",
    y = "Count"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = danceability)) +
  geom_histogram(fill = "darkgreen", bins = 30, color = "white") +
  labs(
    title = "Distribution of Danceability",
    x = "Danceability",
    y = "Count"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = energy)) +
  geom_histogram(fill = "salmon", bins = 30, color = "white") +
  labs(
    title = "Distribution of Energy",
    x = "Energy",
    y = "Count"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = valence)) +
  geom_histogram(fill = "violet", bins = 30, color = "white") +
  labs(
    title = "Distribution of Valence",
    x = "Valence",
    y = "Count"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = tempo)) +
  geom_histogram(fill = "orange", bins = 30, color = "white") +
  labs(
    title = "Distribution of Tempo",
    x = "Tempo (BPM)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
ggplot(spo1, aes(x = danceability, y = popularity)) +
  geom_point(color = "darkgreen", alpha = 0.5) +
  labs(
    title = "Popularity vs Danceability",
    x = "Danceability",
    y = "Popularity"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = energy, y = popularity)) +
  geom_point(color = "salmon", alpha = 0.5) +
  labs(
    title = "Popularity vs Danceability",
    x = "Energy",
    y = "Popularity"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = valence, y = popularity)) +
  geom_point(color = "violet", alpha = 0.5) +
  labs(
    title = "Popularity vs Valence",
    x = "Valence",
    y = "Popularity"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = tempo, y = popularity)) +
  geom_point(color = "orange", alpha = 0.5) +
  labs(
    title = "Popularity vs Tempo",
    x = "Tempo",
    y = "Popularity"
  ) +
  theme_minimal()
```

The diagram is not clear.
```{r}
cor(spo1$danceability, spo1$popularity)
cor(spo1$energy, spo1$popularity)
cor(spo1$valence, spo1$popularity)
cor(spo1$tempo, spo1$popularity)
```
```{r}
spo1$pop_group <- cut(
  spo1$popularity,
  breaks = c(0, 20, 40, 60, 80, 100),
  labels = c("Very Low", "Low", "Medium", "High", "Very High"),
  include.lowest = TRUE
)

spo1$pop_group <- as.factor(spo1$pop_group)
table(spo1$pop_group)
```
```{r}
ggplot(spo1, aes(x = pop_group, y = danceability, fill = pop_group)) +
  geom_boxplot() +
  labs(
    title = "Danceability across Popularity Groups",
    x = "Popularity Group",
    y = "Danceability"
  ) +
  theme_minimal() +
  guides(fill = "none")
ggplot(spo1, aes(x = pop_group, y = energy, fill = pop_group)) +
  geom_boxplot() +
  labs(
    title = "Energy across Energy Groups",
    x = "Popularity Group",
    y = "Energy"
  ) +
  theme_minimal() +
  guides(fill = "none")
ggplot(spo1, aes(x = pop_group, y = valence, fill = pop_group)) +
  geom_boxplot() +
  labs(
    title = "Valence across Popularity Groups",
    x = "Popularity Group",
    y = "Valence"
  ) +
  theme_minimal() +
  guides(fill = "none")
ggplot(spo1, aes(x = pop_group, y = tempo, fill = pop_group)) +
  geom_boxplot() +
  labs(
    title = "Tempo across Popularity Groups",
    x = "Popularity Group",
    y = "Tempo"
  ) +
  theme_minimal() +
  guides(fill = "none")
```
```{r}
anova_dance <- aov(danceability ~ pop_group, data = spo1)
summary(anova_dance)
```
```{r}
get_eta_sq <- function(feature_name, data) {
  formula <- as.formula(paste(feature_name, "~ pop_group"))
  model <- aov(formula, data = data)
  tab <- summary(model)[[1]]
  
  SS_between <- tab["pop_group", "Sum Sq"]
  SS_total   <- sum(tab[, "Sum Sq"])
  
  eta_sq <- SS_between / SS_total
  
  return(list(
    feature = feature_name,
    eta_sq = eta_sq,
    p_value = tab["pop_group", "Pr(>F)"]
  ))
}
results <- rbind(
  unlist(get_eta_sq("danceability", spo1)),
  unlist(get_eta_sq("energy", spo1)),
  unlist(get_eta_sq("valence", spo1)),
  unlist(get_eta_sq("tempo", spo1))
)

results <- as.data.frame(results)
results
```

```{r}
ggplot(spo1,
       aes(x = danceability,
           y = popularity,
           color = family)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Popularity vs Danceability by Music Family",
    x = "Danceability",
    y = "Popularity",
    color = "Family"
  ) +
  theme_minimal()

ggplot(spo1,
       aes(x = energy,
           y = popularity,
           color = family)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Popularity vs Energy by Music Family",
    x = "Energy",
    y = "Popularity",
    color = "Family"
  ) +
  theme_minimal()

ggplot(spo1,
       aes(x = valence,
           y = popularity,
           color = family)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Popularity vs Valence by Music Family",
    x = "Valence",
    y = "Popularity",
    color = "Family"
  ) +
  theme_minimal()

ggplot(spo1,
       aes(x = tempo,
           y = popularity,
           color = family)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Popularity vs Tempo by Music Family",
    x = "Tempo",
    y = "Popularity",
    color = "Family"
  ) +
  theme_minimal()
```

```{r}
anova_cluster <- aov(popularity ~ cluster, data = spo1)
summary(anova_cluster)
tab    <- summary(anova_cluster)[[1]]
eta_sq <- tab["cluster", "Sum Sq"] / sum(tab[, "Sum Sq"])
cat(sprintf("Eta squared (η²) = %.4f\n", eta_sq))
tukey_cluster <- TukeyHSD(anova_cluster)
xkablesummary(tukey_cluster)
```
```{r}
cluster_genre_tab <- table(spo1$cluster, spo1$track_genre)
cluster_genre_tab[1:4, 1:10]
```
```{r}
anova_cluster <- aov(popularity ~ cluster, data = spo1)
summary(anova_cluster)
```

Meg-
Can we see any differences in Genre when we look at Tempo x Popularity? There are too many points! 

```{r}
ggplot(spo1, aes(x = tempo, y = popularity, color= track_genre)) + geom_point(alpha = 0.5)
```
What is another way to look at plots by Genre? Possibly in facets!
```{r}
ggplot(spo1, aes(x = tempo, y = popularity)) + geom_point(color= "blue", alpha = 0.3) + facet_wrap(~ track_genre) 
```

Are explicit tracks more or less popular than non-explicit tracks?
we want to do a t-test to check:
```{r}
t.test(popularity ~ explicit, data = spo1)
```
this returns a p-value <2e-16, which is significant
```{r}
lower <- quantile(spo1$popularity, 0.25, na.rm = TRUE)
upper <- quantile(spo1$popularity, 0.75, na.rm = TRUE)
spo1_top_pop<- subset(spo1, popularity >= upper)
t.test(popularity ~ explicit, data = spo1_top_pop)
spo1_middle_pop <- subset(spo1, popularity > lower & popularity <= upper)
t.test(popularity ~ explicit, data = spo1_middle_pop)
spo1_bottom_pop <- subset(spo1, popularity < lower)
t.test(popularity ~ explicit, data = spo1_bottom_pop)
```
```{r}
cut_pop90<- quantile(spo1$popularity, 0.90)
spo1_pop90 <- subset(spo1, popularity > cut_pop90)
cut_pop95<- quantile(spo1$popularity, 0.95)
spo1_pop95 <- subset(spo1, popularity > cut_pop95)
t.test(popularity ~ explicit, data = spo1_pop90)
t.test(popularity ~ explicit, data = spo1_pop95)
```
#Splitting popularity by values 
```{r}
pop_20df <- spo1 %>%
  filter(popularity >= 0 & popularity <= 20)
pop_40df <- spo1 %>%
  filter(popularity >= 20 & popularity <= 40)
pop_60df <- spo1 %>%
  filter(popularity >= 40 & popularity <= 60)
pop_80df <- spo1 %>%
  filter(popularity >= 60 & popularity <= 80)
pop_100df <- spo1 %>%
  filter(popularity >= 80 & popularity <= 100)
```
#Correlation between explicit and popularity- shows extremely low
```{r}
spo1$explicit_num<- as.numeric(spo1$explicit)
cor(spo1$popularity, spo1$explicit_num)
````

#Calculating correlation values of audio features(danceability, energy, tempo and valence) to popularity for each group

```{r}
features <- c("danceability", "energy", "valence", "tempo", "popularity")

cor_pop_20 <- cor(pop_20df[, features], pop_20df$popularity,
use = "complete.obs")
cor_pop_40 <- cor(pop_40df[, features], pop_40df$popularity,
use = "complete.obs")
cor_pop_60 <- cor(pop_60df[, features], pop_60df$popularity,
use = "complete.obs")
cor_pop_80 <- cor(pop_80df[, features], pop_80df$popularity,
use = "complete.obs")
cor_pop_100 <- cor(pop_100df[, features], pop_100df$popularity,
use = "complete.obs")
```

#Plotting heatmap of correlation values

```{r}
library(reshape2)
library(ggplot2)

plot_corr <- function(cor_matrix, title_text, filename) {
  
  melted <- melt(cor_matrix)
  
  p <- ggplot(melted, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    geom_text(aes(label = round(value, 2)),
              size = 4) +
    scale_fill_gradient2(low = "blue",
                         mid = "white",
                         high = "red",
                         midpoint = 0,
                         limits = c(-1,1)) +
    theme_minimal() +
    labs(title = title_text,
         x = "",
         y = "",
         fill = "Correlation")
  
  ggsave(filename, plot = p, width = 6, height = 5, dpi = 300)
  
  return(p)
}


plot_corr(cor_pop_20,
          "Correlation Matrix: Popularity 0-20 Range", "corr_pop_20.png")
plot_corr(cor_pop_40,
          "Correlation Matrix: Popularity 20-40 Range", "corr_pop_40.png")
plot_corr(cor_pop_60,
          "Correlation Matrix: Popularity 40-60 Range", "corr_pop_60.png")
plot_corr(cor_pop_80,
          "Correlation Matrix: Popularity 60-80 Range", "corr_pop_80.png")
plot_corr(cor_pop_100,
          "Correlation Matrix: Popularity 80-100 Range", "corr_pop_100.png")
```

#Converting correlation values from wide format into the long format to compare each audio feature across different popularity groups
```{r}
features <- c("danceability", "energy", "valence", "tempo")

df_list <- list(
  "0-20"   = pop_20df,
  "20-40"  = pop_40df,
  "40-60"  = pop_60df,
  "60-80"  = pop_80df,
  "80-100" = pop_100df
)

cor_long <- imap_dfr(df_list, ~{
  d <- .x
  tibble(
    Pop_Group = .y,
    Feature = features,
    Correlation = map_dbl(features, \(f) cor(d[[f]], d$popularity, use = "complete.obs"))
  )
})

cor_long$Pop_Group <- factor(cor_long$Pop_Group,
                             levels = c("0-20","20-40","40-60","60-80","80-100"))
```

#Plotting a line plot for correlation values between danceability and popularity across different popularity groups
```{r}
dance_cor <- cor_long %>% filter(Feature == "danceability")

p <- ggplot(dance_cor, aes(x = Correlation , y = Pop_Group, group = 1)) +
  geom_line() +
  geom_point(size = 3, color = "firebrick") +
  theme_minimal() +
  labs(title = "Correlation(popularity, danceability) across popularity groups",
       x = "Correlation", y = "Popularity group")

p
ggsave("danceability_corr_across_groups.png", p, width = 7, height = 4.5, dpi = 300)
```

#Plotting a line plot for correlation values between energy and popularity across different popularity groups

```{r}
energy_cor <- cor_long %>% filter(Feature == "energy")

p <- ggplot(energy_cor, aes(x = Correlation , y = Pop_Group, group = 1)) +
  geom_line() +
  geom_point(size = 3, color = "steelblue") +
  theme_minimal() +
  labs(title = "Correlation(popularity, energy) across popularity groups",
       x = "Correlation", y = "Popularity group")

p
ggsave("energy_corr_across_groups.png", p, width = 7, height = 4.5, dpi = 300)

```

#Plotting a line plot for correlation values between valence and popularity across different popularity groups

```{r}
valence_cor <- cor_long %>% filter(Feature == "valence")

p <- ggplot(valence_cor, aes(x = Correlation , y = Pop_Group, group = 1)) +
  geom_line() +
  geom_point(size = 3, color = "aquamarine") +
  theme_minimal() +
  labs(title = "Correlation(popularity, valence) across popularity groups",
       x = "Correlation", y = "Popularity group")

p
ggsave("valence_corr_across_groups.png", p, width = 7, height = 4.5, dpi = 300)
```

#Plotting a line plot for correaltion values between tempo and popularity across different popularity groups

```{r}
tempo_cor <- cor_long %>% filter(Feature == "tempo")

p <- ggplot(tempo_cor, aes(x = Correlation , y = Pop_Group, group = 1)) +
  geom_line() +
  geom_point(size = 3, color = "coral") +
  theme_minimal() +
  labs(title = "Correlation(popularity, tempo) across popularity groups",
       x = "Correlation", y = "Popularity group")

p
ggsave("tempo_corr_across_groups.png", p, width = 7, height = 4.5, dpi = 300)

```
