---
title: "HW 6: Hypothesis Tests"
author: "Ruofei(Sophia) Du, Meghan Phillips, Harivinay Gujjula"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
```


```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(results="markup", warning = F, message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

```{r}
spo <- read.csv("spotifydataset.csv")
str(spo)
summary(spo)
dim(spo)
```
Clean the data and get rid of the columns we don't need.
```{r}
spo1 <- spo[, c(
  "popularity",
  "danceability",
  "energy",
  "valence",
  "tempo",
  "explicit",
  "track_genre"
)]
str(spo1)
summary(spo1)
colSums(is.na(spo1))
```
```{r}
spo1$explicit <- as.factor(spo1$explicit)
str(spo1)
```
Furthur cleaning the data, as the BPM can't be 0, but the summary showed that the min(tempo) = 0.
```{r}
sum(spo1$tempo == 0)
```
We can see that there are still 157 rows with tempo = 0, which is inproper cause we can't have a song with BPM = 0. We get rid of those missing data.

```{r}
spo1 <- spo1[spo1$tempo > 0, ]
sum(spo1$tempo == 0)
write.csv(spo1,"cleaned_spodata.csv")
```

Now let's start with EDA. First, we plot some ggplots to observe the distribution. 
```{r}
library(ggplot2)
ggplot(spo1, aes(x = popularity)) +
  geom_histogram(fill = "skyblue", bins = 30, color = "white") +
  labs(
    title = "Distribution of Track Popularity",
    x = "Popularity Score",
    y = "Count"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = danceability)) +
  geom_histogram(fill = "darkgreen", bins = 30, color = "white") +
  labs(
    title = "Distribution of Danceability",
    x = "Danceability",
    y = "Count"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = energy)) +
  geom_histogram(fill = "salmon", bins = 30, color = "white") +
  labs(
    title = "Distribution of Energy",
    x = "Energy",
    y = "Count"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = valence)) +
  geom_histogram(fill = "violet", bins = 30, color = "white") +
  labs(
    title = "Distribution of Valence",
    x = "Valence",
    y = "Count"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = tempo)) +
  geom_histogram(fill = "orange", bins = 30, color = "white") +
  labs(
    title = "Distribution of Tempo",
    x = "Tempo (BPM)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
ggplot(spo1, aes(x = danceability, y = popularity)) +
  geom_point(color = "darkgreen", alpha = 0.5) +
  labs(
    title = "Popularity vs Danceability",
    x = "Danceability",
    y = "Popularity"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = energy, y = popularity)) +
  geom_point(color = "salmon", alpha = 0.5) +
  labs(
    title = "Popularity vs Danceability",
    x = "Energy",
    y = "Popularity"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = valence, y = popularity)) +
  geom_point(color = "violet", alpha = 0.5) +
  labs(
    title = "Popularity vs Valence",
    x = "Valence",
    y = "Popularity"
  ) +
  theme_minimal()
```
```{r}
ggplot(spo1, aes(x = tempo, y = popularity)) +
  geom_point(color = "orange", alpha = 0.5) +
  labs(
    title = "Popularity vs Tempo",
    x = "Tempo",
    y = "Popularity"
  ) +
  theme_minimal()
```

The diagram is not clear, maybe we should do some sampling. 

Meg-
Can we see any differences in Genre when we look at Tempo x Popularity? There are too many points! 

```{r}
ggplot(spo1, aes(x = tempo, y = popularity, color= track_genre)) + geom_point(alpha = 0.5)
```
What is another way to look at plots by Genre? Possibly in facets!
```{r}
ggplot(spo1, aes(x = tempo, y = popularity)) + geom_point(color= "blue", alpha = 0.3) + facet_wrap(~ track_genre) 
```

Are explicit tracks more or less popular than non-explicit tracks?
we want to do a t-test to check:
```{r}
t.test(popularity ~ explicit, data = spo1)
```
this returns a p-value <2e-16, which is significant
```{r}
lower <- quantile(spo1$popularity, 0.25, na.rm = TRUE)
upper <- quantile(spo1$popularity, 0.75, na.rm = TRUE)
spo1_top_pop<- subset(spo1, popularity >= upper)
t.test(popularity ~ explicit, data = spo1_top_pop)
spo1_middle_pop <- subset(spo1, popularity > lower & popularity <= upper)
t.test(popularity ~ explicit, data = spo1_middle_pop)
spo1_bottom_pop <- subset(spo1, popularity < lower)
t.test(popularity ~ explicit, data = spo1_bottom_pop)
```
```{r}
cut_pop90<- quantile(spo1$popularity, 0.90)
spo1_pop90 <- subset(spo1, popularity > cut_pop90)
cut_pop95<- quantile(spo1$popularity, 0.95)
spo1_pop95 <- subset(spo1, popularity > cut_pop95)
t.test(popularity ~ explicit, data = spo1_pop90)
t.test(popularity ~ explicit, data = spo1_pop95)
```
